Icicle: write once, run once

### 2016-05-28 2016fhpc/section/01-Introduction.tex
some changes to introduction, s2, s3

At Ambiata, we peform feature generation for machine learning applications by executing many thousands of simple queries over hundreds of gigabytes worth of data.\footnote{In 2016 this was a lot of data.} For such applications we must automatically fuse these separate queries and be sure that the result can be executed in a single pass over the input. We also ingest multiple gigabytes of new data per day, and must incrementally update existing features without recomputing them all from scratch. Our feature generation process is executed in parallel on hundreds of nodes on a cloud based system, and if we performed neither fusion or incremental update then the cost of the computation would begin to exceed the salaries of the developers.

@@

\item
We present an fold-based intermediate language, which allows the query fusion transformation to be a simple matter of appending two intermediate programs\sref{s:IcicleCore};

\item
We present production benchmarks of Icicle compiled code, which compare favourably against standard unix utilities, and outperform an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
\end{itemize}

>>>>

At Ambiata, we peform feature generation for machine learning applications by executing many thousands of simple queries over terabytes worth of compressed data.\footnote{In 2016 this was a lot of data.}
For such applications we must automatically fuse these separate queries and be sure that the result can be executed in a single pass over the input.
We also ingest tens of gigabytes of new data per day, and must incrementally update existing features without recomputing them all from scratch.
Our feature generation process is executed in parallel on hundreds of nodes on a cloud based system, and if we performed neither fusion or incremental update then the cost of the computation would begin to exceed the salaries of the developers.

@@

\item
We present a fold-based intermediate language, which allows the query fusion transformation to be a simple matter of appending two intermediate programs\sref{s:IcicleCore};

\item
We present production benchmarks of Icicle compiled code on our smaller single-disk datasets, which compare favourably against standard unix utilities, and outperform an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
\end{itemize}

### 2016-05-28 2016fhpc/section/02-Source.tex
some changes to introduction, s2, s3

To allow incremental computation, all Icicle queries must execute in a single pass over the input stream. Of course, not all queries can be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in an order that is different to what a stream can provide. However, as we saw in the introduction, although it my not be possible to evaluate a particular \emph{algorithm} in an streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm. Here is the unstreamable example again from the introduction:
\begin{code}
  table kvs { key : Date; value : Real }
  query avg = let k = last key
              in  filter (key == k) of mean value;
\end{code}

@@

The trick to compute our average in an incremental fashion is to recognise that the @filter@ operation is selecting a particular sub-set of values from the input, but the value computed from this sub-set depends only on the values in that sub-set, and no other information. Instead of computing the mean of a single subset whose identity is only known at the end of the stream, we can instead compute the mean of \emph{all possible subsets}, and return the required one once we know what that is:
\begin{code}
  table kvs { key : Date; value : Real } 
  query avg = let k    = last  key in
              let avgs = group key of mean value
              in  lookup k avgs
\end{code}

@@

@@

The expression takes the input stream @xs@ and selects the elements greater than zero, and pairs them with elements that are less than zero. This expression requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then the positive values must be retained before any output can be produced.

The language constructs of Icicle are designed so that queries that would require unbounded buffering cannot be written, with one major caveat that we will discuss in a moment. In Icicle, the stream being processed (such as @xs@ above) is implicit in the defined query, and is never named explicitly. Constructs such as @filter@ and @fold@ do not take the name of the stream as argument, but instead operate on stream defined in the context. In all the example Icicle queries so far, the name of the table / stream is not mentioned in the body of the query, and serves as a comment only. Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query. In the expression @filter p of mean value@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.

Finally, note that our major caveat is that @group@ combinator we used in \S2 uses space proportional to the number of distinct \emph{keys} in the input stream. For our applications the keys are commonly company names, customer names, and days of the year. Our production system knows that these types are bounded in size and that maps from keys to values will fit easily in memory. Attempting to group by values of a type with an unbounded number of members, such as a @Real@ or @String@ results in a compile-time warning.

@@

The grammar for Icicle is given in figure~\ref{fig:source:grammar}.
Value types are given in $T$ and can be numbers, booleans or maps.
Modal types, $\TauMode$ can be a pure value type, or a modality associated with a value type.
Function types, $\TauFun$, can be either a non-function modal type, or a function from modal type to function type.
As Icicle is a first-order language, function types are not value types.

The table definition in $\mi{Table}$ gives a table name and the names and types of columns.
$\mi{Exp}$ defines the expressions used as query and function bodies.
$\mi{Prim}$ defines the primitive types.
Function and query definitions are given in $\mi{Def}$.
$\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.

Expressions can be variable names, primitive operators such as addition and division, function application, let bindings, or query operations.
The query operations can define a fold such as a sum or a count, filtering according to a predicate, or grouping by a key.
Function application $\mi{x~Exp*}$ is the name of a top-level function applied to any number of arguments.
Primitive application $\mi{Prim~Exp*}$ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.

@@

The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.

@@

Filters take a predicate, which can be thought of as a stream of booleans, and a query to perform for the satisfying values.

@@

Group takes a key to group by, and a query to perform on the values of each grouping.

@@

$\mi{Def}$ is used for function and query definitions.
% The @count@ function takes no arguments and returns a fold counting the number of elements.
% \begin{code}
% function count
%  = fold c = 0 then c + 1;
% \end{code}

@@

Finally, $\mi{Top}$ puts all these together with a table definition, function definitions, and query definitions in one place.
All queries here operate over the same input table, and will be fused into a single loop over the input data.

>>>>

To allow incremental computation, all Icicle queries must execute in a single pass over the input stream.
Of course, not all queries can be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in a different order to what the stream provides.
However, as we saw in the introduction, although a particular \emph{algorithm} may be impossible to evaluate in a streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm.
Here is the unstreamable example again from the introduction:
\begin{code}
  table kvs { key : Date; value : Real }
  query avg = let k = last key
              in  filter (key == k) of mean value;
\end{code}

@@

The trick to compute our average in an incremental fashion is to recognise that the @filter@ operation is selecting a particular subset of values from the input, but the value computed from this subset depends only on the values in that subset, and no other information. Instead of computing the mean of a single subset whose identity is only known at the end of the stream, we can instead compute the mean of \emph{all possible subsets}, and return the required one once we know what that is:
\begin{code}
  table kvs { key : Date; value : Real } 
  query avg = let k    = last  key in
              let avgs = group key of mean value
              in  lookup k avgs
\end{code}

@@

When exposing impure computations whose value changes across successive runs, these must be bound as aggregates.
For example, we can bind the current time @now@ as an @Aggregate Date@.
This ensures that reused parts of an incremental computation are the same regardless of which day they are executed.

@@

The expression takes the input stream @xs@ and selects the elements greater than zero, and pairs them with elements that are less than zero. This expression requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then all the positive values must be retained before any output can be produced.

The language constructs of Icicle are designed so that queries that would require unbounded buffering cannot be written, with one major caveat that we will discuss in a moment. In Icicle, the stream being processed (such as @xs@ above) is implicit in the defined query, and is never named explicitly. Constructs such as @filter@ and @fold@ do not take the name of the stream as argument, but instead operate on the stream defined in the context. In all the example Icicle queries so far, the name of the table / stream is not mentioned in the body of the query, and serves as a comment only. Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query. In the expression @filter p of mean value@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.

Finally, note that our major caveat is that the @group@ combinator we used in \S2 uses space proportional to the number of distinct \emph{keys} in the input stream.
For our applications the keys are commonly company names, customer names, and days of the year.
Grouping by types with an unbounded number of members, such as @Real@ or @String@ can be undesirable, and we wish to outlaw this in a future version of our production compiler.
% Our production system knows that these types are bounded in size and that maps from keys to values will fit easily in memory. Attempting to group by values of a type with an unbounded number of members, such as a @Real@ or @String@ results in a compile-time warning.

@@

The grammar for Icicle is given in figure~\ref{fig:source:grammar}.
Value types, $T$, can be numbers, booleans or maps.
Modal types, $\TauMode$ can be a pure value type, or a modality associated with a value type.
Function types, $\TauFun$, are a list of argument types, and a single return type.
As Icicle is a first-order language, function types are not value types.

$\mi{Table}$ defines a table with a name and the names and types of columns.
$\mi{Exp}$ defines the expressions used as query and function bodies.
Primitive application is defined as prefix, but we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.
$\mi{Prim}$ defines the primitive types.
Function and query definitions are given in $\mi{Def}$.
$\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.
All queries in a $\mi{Top}$ are fused into a single loop over the input data.

We now take a more detailed look at the group example from \S2, as these versions did not define the functions @last@ and @mean@.
We write a function @count@, which takes no arguments.
Its definition is a fold that starts at zero, and increments by one for every element.
\begin{code}
  function count
   = fold c = 0 then c + 1;
\end{code}

Next, @sum@ takes one argument, which is the elements to compute the sum of.
Its definition is a fold that starts at zero, and increments by the sum element for each element.
We can then define @mean@ as the sum divided by the count.
\begin{code}
  function sum  (v : Element Real)
   = fold s = 0 then s + v;
  function mean (v : Element Real)
   = sum v / count;
\end{code}

Finally, the @last@ function takes the elements of keys, and for every element returns that element.
Here we use an arbitrary initial value of 1900-01-01, but in reality this would return an optional value.
\begin{code}
  function last (k : Element Date)
   = fold l = date 1900 1 1 then k;
\end{code}

% Expressions can be variable names, primitive operators such as addition and division, function application, let bindings, or query operations.
% The query operations can define a fold such as a sum or a count, filtering according to a predicate, or grouping by a key.
% Function application $\mi{x~Exp*}$ is the name of a top-level function applied to any number of arguments.
% Primitive application $\mi{Prim~Exp*}$ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.

@@

% The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.

@@

% Filters take a predicate, which can be thought of as a stream of booleans, and a query to perform for the satisfying values.

@@

% Group takes a key to group by, and a query to perform on the values of each grouping.

@@

% $\mi{Def}$ is used for function and query definitions.
% The @count@ function takes no arguments and returns a fold counting the number of elements.
% \begin{code}
% function count
%  = fold c = 0 then c + 1;
% \end{code}

@@

% Finally, $\mi{Top}$ puts all these together with a table definition, function definitions, and query definitions in one place.
% All queries here operate over the same input table, and will be fused into a single loop over the input data.
